An Empirical Analysis of the Role of Energy in Economic Growth

### 2013-02-22 EconGrowth.Rnw
Now creating a CES comparison graph.

<<ces functions, eval=TRUE>>=
######################################
# Loads data and trims rows where necessary.
# Also, replaces the name of the desired energy column with "iEToFit".
# Returns a data.frame with the loaded data.
##
loadAndPrepDataForCES <- function(countryAbbrev, energyType){
  dataTable <- loadData(countryAbbrev)
  if (energyType == "U"){
    # Trim the dataset to include only those years for which U is available.
    dataTable <- subset(dataTable, !is.na(iU))
  }
  # Trim data from China and South Africa that is missing labor information for 2011.
  dataTable <- subset(dataTable, !is.na(iLabor))
  # We need to do the CES fit with the desired energyType.
  # To achieve the correct fit, we'll change the name of the desired column
  # to "iEToFit" and use "iEToFit" in the nls function.
  dataTable <- replaceColName(dataTable, energyType, "iEToFit")
  return(dataTable)
}

####################
# Returns an nls ces model for the country and energyType specified.
# energyType should be one of Q", "X", or "U".
##
cesModel <- function(countryAbbrev, energyType){
  # Load the data that we need.
  dataTable <- loadAndPrepDataForCES(countryAbbrev, energyType)
#   # Establish default guesses for sigma, gamma_E, beta, lambda_L and lambda_E.
#   start <- list(sigma=0.9, gamma_E=0.01, beta=0.7, lambda_L=0.011, lambda_E=0.018) #values for Sweden
#   # Establish bounds
#   lower = list(sigma=-Inf, gamma_E=0.0, beta=-Inf, lambda_L=-Inf, lambda_E=-Inf)
#   upper = list(sigma= Inf, gamma_E=1.0, beta= Inf, lambda_L= Inf, lambda_E= Inf)
#   # Now some country-specific values.
#   if (countryAbbrev == "US"){
#     start <- list(sigma=0.95, gamma_E=0.1, beta=0.7, lambda_L=0.015, lambda_E=0.015) #values for Sweden
#   }
#   if (countryAbbrev == "UK"){
#     start <- list(sigma=0.95, gamma_E=0.2, beta=0.7, lambda_L=0.02, lambda_E=0.02) #values for Sweden
#     lower = list(sigma=-Inf, gamma_E=0.0, beta=-Inf, lambda_L=-0.3, lambda_E=-0.3)
#     upper = list(sigma= Inf, gamma_E=1.0, beta= Inf, lambda_L= 0.3, lambda_E= 0.3)
#   }
#   if (countryAbbrev == "CN"){
#     
#   }
#   # Runs a non-linear least squares fit to the data with constraints
#   # This is the version of the function from the Stern and Kander (2012) paper. However, it includes a 
#   # 1/sigma exponent on the gamma_E terms, which, I believe, is bogus.
# #   model <- iGDP ~ ((1-gamma_E)^(1.0/sigma) * ((exp(lambda_L*iYear))^beta 
# #                                                   * iLabor^beta
# #                                                   * iCapStk^(1-beta) )^((sigma-1.0)/sigma) 
# #                   + gamma_E^(1.0/sigma)*(exp(lambda_E*iYear) * iEToFit)^((sigma-1.0)/sigma))^(sigma/(sigma-1.0))
#   
#   # This is the corrected version of the Stern and Kander (2012) paper.
# #   model <- iGDP ~ ((1-gamma_E) * ((exp(lambda_L*iYear))^beta * iLabor^beta * iCapStk^(1-beta) )^((sigma-1.0)/sigma) 
# #                   + gamma_E * (exp(lambda_E*iYear) * iEToFit)^((sigma-1.0)/sigma))^(sigma/(sigma-1.0))
#   
#   # This is the nls fit to the CES model  
# #   modelCES <- nls(formula=model, data=dataTable, start=start,
# #                   control=nls.control(maxiter=10000, tol=1e-05, minFactor=1/1024, 
# #                                       printEval=TRUE, warnOnly=TRUE),
# #                   algorithm = "port",
# #                   lower = lower,
# #                   upper = upper,
# #                   trace=TRUE
# #                   )
# 
#   # An attempt with nlmrt function nlxb. Doesn't provide many good results.
# #   modelCES <- nlxb(model,
# #                    data=dataTable,
# #                    start=list(sigma=sigmaGuess, gamma_E=gamma_EGuess, beta=betaGuess, 
# #                               lambda_L=lambda_LGuess, lambda_E=lambda_EGuess),
# #                    trace=TRUE
# #                   )

  # Try a fit to a nested CES function with the micEconCES package.  
  # This approach is based on a white paper by Henningsen and Henningsen 
  # http://okonomi.foi.dk/workingpapers/WPpdf/WP2011/WP_2011_9_econometric_estimation_CES.pdf.
  # Set a few of the variables up front
  yName = "iGDP"
  xNames = c("iCapStk", "iLabor", "iEToFit")
  tName = "iYear"

@@

# print("US U")
# print(summary(cesModel("US", "U")))
# print("UK U")
# print(summary(cesModel("UK", "U")))
# print("JP U")
# print(summary(cesModel("JP", "U")))

# print("")
# print("CN Q")
# print(summary(cesModel("CN", "Q")))
# 
# print("")
# print("ZA Q")
# print(summary(cesModel("ZA", "Q")))
# 
# print("")
# print("SA Q")
# print(summary(cesModel("SA", "Q")))
# 
# print("")
# print("IR Q")
# print(summary(cesModel("IR", "Q")))
# 
# print("")
# print("TZ Q")
# print(summary(cesModel("TZ", "Q")))
# 
# print("")
# print("ZM Q")
# print(summary(cesModel("ZM", "Q")))

#########################
# Takes the CES fitted models and creates per-country predictions for them.
# Returns a data.frame with the predictions.
##
cesPredictions <- function(countryAbbrev, energyType){
  xNames = c("iCapStk", "iLabor", "iEToFit")
  tName = "iYear"
  dataTable <- loadAndPrepDataForCES(countryAbbrev, energyType)
  # Can't make predictions for any of CN, ZA, SA, IR, TZ, or ZM if we're interested in U
  if (!(haveDataSF(countryAbbrev, energyType))){
    # If we don't have data for this combination of countryAbbrev and energyType, 
    # return a column of NAs if the above conditions have been met.
    nRows <- 21 # All of these countries need 21 rows.
    df <- as.data.frame(matrix(NA, ncol = 1, nrow = nRows))
    colnames(df) <- "pred"
    return(df)
  }
  model <- cesModel(countryAbbrev, energyType)
  print(model)
  pred <- cesCalc(xNames=xNames, data=dataTable, coef=coef(model), tName="iYear", nested=TRUE)
  df <- data.frame(pred)
  # Pad with rows as necessary
  df <- padRows(countryAbbrev, df)
  return(df)
}

print(cesPredictions("US", "Q"))

@@

  # Now create the data for a table.
  dataCES <- rbind(upperCES, estCES, lowerCES)
  print(dataCES)
  return(dataCES)
}

@@

<<CESq analysisQ, eval=FALSE>>=
countryName <- "US"
energyType <- "Q"
tableCESq <- cesTable(countryName, energyType)

#CESqTables <- lapply(countryAbbrevs, cesTable, energyType="Q")
@

<<CESq Results, results='asis', eval=FALSE>>=
print(tableCESq, caption.placement="top")

#print(CESqTables[["US"]], caption.placement="top")
#print(CESqTables[["ZA"]], caption.placement="top")
@

@@

@@

Figure \ref{fig:LINEX_GDP_Lattice_Graph} compares LINEX model predictions with historical data.

>>>>

<<ces functions, eval=TRUE>>=
# Set some common names for CES variables.
xNames = c("iCapStk", "iLabor", "iEToFit")
tName = "iYear"
yName = "iGDP"

######################################
# Loads data and trims rows where necessary.
# Also, replaces the name of the desired energy column with "iEToFit".
# Returns a data.frame with the loaded data.
##
loadAndPrepDataForCES <- function(countryAbbrev, energyType){
  dataTable <- loadData(countryAbbrev)
  if (energyType == "U"){
    # Trim the dataset to include only those years for which U is available.
    dataTable <- subset(dataTable, !is.na(iU))
  }
  # Trim data from China and South Africa that is missing labor information for 2011.
  dataTable <- subset(dataTable, !is.na(iLabor))
  # We need to do the CES fit with the desired energyType.
  # To achieve the correct fit, we'll change the name of the desired column
  # to "iEToFit" and use "iEToFit" in the nls function.
  dataTable <- replaceColName(dataTable, energyType, "iEToFit")
  return(dataTable)
}

####################
# Returns an nls ces model for the country and energyType specified.
# energyType should be one of Q", "X", or "U".
cesModel <- function(countryAbbrev, energyType){
  # Load the data that we need.
  dataTable <- loadAndPrepDataForCES(countryAbbrev, energyType)

@@

#########################
# Takes the CES fitted models and creates per-country predictions for them.
# Returns a data.frame with the predictions.
##
cesPredictions <- function(countryAbbrev, energyType){
  dataTable <- loadAndPrepDataForCES(countryAbbrev, energyType)
  # Can't make predictions for any of CN, ZA, SA, IR, TZ, or ZM if we're interested in U
  if (!(haveDataSF(countryAbbrev, energyType))){
    # If we don't have data for this combination of countryAbbrev and energyType, 
    # return a column of NAs if the above conditions have been met.
    nRows <- 21 # All of these countries need 21 rows.
    df <- as.data.frame(matrix(NA, ncol = 1, nrow = nRows))
    colnames(df) <- "pred"
    return(df)
  }
  model <- cesModel(countryAbbrev, energyType)
  pred <- cesCalc(xNames=xNames, data=dataTable, coef=coef(model), tName="iYear", nested=TRUE)
  df <- data.frame(pred)
  # Pad with rows as necessary
  df <- padRows(countryAbbrev, df)
  return(df)
}

#########################
# Takes the CES fitted models and creates a single column of predicted GDP values
# that corresponds, row for row, with the AllData.txt file.
##
cesPredictionsColumn <- function(energyType){
  out <- do.call("rbind", lapply(countryAbbrevs, cesPredictions, energyType=energyType))
  colnames(out) <- c(paste("predGDP", energyType, sep=""))
  return(out)
}

##############################
# Creates a graph that plots predicted GDP as lines, one for each single factor, and historical GDP 
# data as open circles.
##
createCESLatticeGraph <- function(){
  dataTable <- loadData("All") #Grab the raw data
  predictionsQ <- cesPredictionsColumn("Q") #Predictions from CES with Q
  predictionsX <- cesPredictionsColumn("X") #Predictions from CES with X
  predictionsU <- cesPredictionsColumn("U") #Predictions from CES with U
  #Now add the predictions columns to the data.
  dataTable <- cbind(dataTable, predictionsQ, predictionsX, predictionsU) 
  graphType <- "b" #b is for both line and symbol
  lineTypes <- c(0, 2, 4, 1) #line types. See http://en.wikibooks.org/wiki/R_Programming/Graphics
  lineWidths <- c(0, 1, 1, 1) #line widths. 0 means no line.
  colors <- c("black", "red", "blue", "darkorange") #line and symbol colors
  symbols <- c(1, NA, NA, NA)  #NA gives no symbol.
  graph <- xyplot(iGDP+predGDPQ+predGDPX+predGDPU ~ Year | Country, data=dataTable,
                  type=graphType,
                  index.cond=list(countryOrderForGraphs), #orders the panels.
                  layout=ninePanelLayoutSpec, #indicates a 3x3 arrangement of panels.
                  strip = strip.custom(factor.levels=countryNamesAlph, bg="white"), #Sets strip parameters
                  as.table=TRUE, #indexing of panels starts in upper left and goes across rows.
                  lty = lineTypes, lwd = lineWidths, col = colors, #Controls line parameters
                  pch = symbols, col.symbol = colors, #Controls symbol parameters
                  key=list(text=list(c("Actual", "With $q$", "With $x$", "With $u$")),
                           type=graphType,
                           cex=keyTextSize, #controls size of text in the key
                           lines=list(lty=lineTypes, lwd=lineWidths), #controls line types
                           pch=symbols, col=colors, #controls symbol (plot characters) types
                           columns=keyColumns, x=keyXLoc, y=keyYLoc), #controls columns and position of the key
                  scales=list(cex=scaleTextSize, #controls text size on scales.
                              tck=scaleTickSize, #controls tick mark length. < 0 for inside the graph.
                              y=list(relation="free", #allows each axis to be different
                                     at=c(1,2,3,4,5,6,7,8,9,10) #tick mark locations
                                     )
                              ), 
                  ylim=list(c(1,8), c(1,3), c(1,3), c(1,3), c(1,3), c(1,3), c(1,3), c(1,3), c(1,3)),
                  xlab="", ylab="Indexed GDP (1980=1 or 1991=1)") #axis labels
  return(graph)
}

@@

  # Now create the data for a table.
  dataCES <- rbind(upperCES, estCES, lowerCES)
  return(dataCES)
}

@@

@@

\subsection{CES comparisons}

Figure \ref{fig:CES_GDP_Lattice_Graph} compares CES model predictions to historical data.

<<CES_GDP_Lattice_Graph, fig.width=ninePanelGraphWidth, fig.height=ninePanelGraphHeight, fig.pos="H", fig.cap="CES results.", eval=TRUE>>=
createCESLatticeGraph()
@

@@

Figure \ref{fig:LINEX_GDP_Lattice_Graph} compares LINEX model predictions to historical data.
